---
title: "1. Data Preparation"
format: html
execute:
  echo: false      # hide code
  warning: false    # hide warnings
  message: false    # hide messages (e.g. package load chatter)
---

```{r}
#| include: false
knitr::opts_chunk$set(results = "hide")
```

```{r}
# Setup
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(sf)

source(here::here("R", "00-settings.R"))

# @TODO - Add some kind of CRS protection to make sure that with any of these spatial operations/conversions we are making the correct assumptions about CRS

```

### 1. Data preparation

* Using placeholder statistically downscaled dataset for bottom temperature, and matched it to the survey grid (prediction grid). Median bottom temperature in June.

* Combine temperature data with the dogfish survey records

* Visual check: map dogfish observations and temperature patterns by year for the full survey and for one region (WCVI)

```{r}
# Prepare dogfish data, including temperature
dog0 <- dogfish # from sdmTMB package
dog <- dog0 |> mutate(log_depth = log(depth))

# `temp_dat0` holds columns for 2002-2099 across climate scenarios (only 1 right now)
# Choose years to extract temp data in
# covariate_years <- min(dog$year):max(dog$year) # historical only
covariate_years <- min(dog$year):2060 # Let's just do some forecasting to 2060 for now, could go to 2099

# Trond dataset - processed by Sean:
# - extracted data for 2002-2099 for one climate scenario for one month (june) and the 50th percentile value only, that works out to a ~70 mb .tif file
# - interpolated it onto the synoptic grid using terra::extract() given the survey grid centroids
# - 1 climate scenario at the median across the model ensemble
temp_dat0 <- readRDS(file.path(data_raw_dir, "thetao_ensemble_sd+ba_bottom_perc_ssp126_p50_survey_blocks_active.rds"))

grid_vect <- terra::vect(temp_dat0) # Convert temp_dat0 to terra SpatVector to be able to use `terra::extract()`
empty_raster <- terra::rast(grid_vect, resolution = 2000) # Empty array that matches dim of temp_dat0

# Subset temp_dat0 to only historical years (for efficiency)
value_cols <- paste0("value_", covariate_years, "_06") # (06 = June)
temp_stack <- terra::rast(lapply(value_cols, \(v) {
  terra::rasterize(grid_vect, empty_raster, field = v)
}))
names(temp_stack) <- covariate_years

dog_vect <- terra::vect(dog, geom = c("longitude", "latitude"), crs = "EPSG:4326") |>
  terra::project(terra::crs(temp_stack))

dog$bottom_temp <- terra::extract(
  temp_stack,
  dog_vect,
  method = "bilinear",
  layer = as.character(dog$year)
)$value

glimpse(dog)

dog_sf <- dog |> sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

```{r}
#| label: wcvi-survey-set-bottom-temp

ggplot(dog_sf) +
  geom_sf(data = pacea::bc_coast) +
  geom_sf(aes(colour = bottom_temp)) +
  viridis::scale_colour_viridis(option = "plasma") +
  facet_wrap(~ year) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
  theme(legend.position = "bottom") +
  ggtitle("Survey set bottom temperatures") +
  viridis::scale_colour_viridis(
    option = "plasma",
    name = "Bottom temperature (°C)",
    guide = guide_colourbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.4, "cm"),
      title.vjust = 0.9,
      direction = "horizontal"
    )
  ) +
  gfplot::coord_sf_auto(dog_sf, buffer = 10000)
```

```{r}
# Save covariate data to use when building prediction grids
# - dp before we filter to WCVI, so we can scale up later
# - keep `cell` (terra's integer cell index) alongside X/Y - temp_df and
#   depth_df are built from separate as.data.frame() calls, and joining on
#   the X/Y doubles later silently drops rows where the two don't compare
#   exactly equal. `cell` is an exact integer key since both rasters share
#   the same `empty_raster` template.
# TODO: will need to replace with more complete depth layer;
depth_rast <- terra::rasterize(grid_vect, empty_raster, field = "depth_m")
survey_rast <- terra::rasterize(grid_vect, empty_raster, field = "survey_abbrev")

depth_df0 <- terra::as.data.frame(c(depth_rast, survey_rast), xy = TRUE, cells = TRUE) |>
  mutate(X = x / 1000, Y = y / 1000, log_depth = log(depth_m)) |>
  select(cell, X, Y, depth_m, log_depth, survey_abbrev) |>
  filter(survey_abbrev == "SYN WCVI")

# survey_abbrev doesn't vary by year, so attach it via a left_join on `cell`
# (not filtering anything out here - just enriching every temp_df row so it
# can be filtered to a region the same simple way as depth_df downstream).
temp_df0 <- terra::as.data.frame(temp_stack, xy = TRUE, cells = TRUE) |>
  tidyr::pivot_longer(-c(cell, x, y), names_to = "year", values_to = "bottom_temp") |>
  mutate(year = as.integer(year), X = x / 1000, Y = y / 1000) |>
  left_join(depth_df0 |> select(cell, survey_abbrev), by = "cell") |>
  select(cell, X, Y, year, bottom_temp, survey_abbrev) |>
  filter(survey_abbrev == "SYN WCVI")
```

```{r}
#| label: trond-bottom-temp

# crop for faster plotting of WCVI only data
wcvi_vect <- terra::vect(temp_dat0[temp_dat0$survey_abbrev == "SYN WCVI", ])
wcvi_temp <- terra::crop(temp_stack, wcvi_vect, mask = TRUE)

ggplot() +
  tidyterra::geom_spatraster(data = wcvi_temp) +
  geom_sf(data = pacea::bc_coast) +
  facet_wrap(~lyr) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
  theme(legend.position = "bottom") +
  ggtitle("Oceanographic data: Median annual June bottom temperature") +
  viridis::scale_fill_viridis(
    option = "plasma",
    name = "Bottom temperature (°C)",
    na.value = "transparent",
    guide = guide_colourbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.4, "cm"),
      # title.position = "top",
      title.vjust = 0.9,
      direction = "horizontal"
    )
  ) +
  gfplot::coord_sf_auto(dog_sf, buffer = 10000) # zoom into WCVI
```

### Split WCVI into two example regions

* Toy N/S split of the WCVI survey grid to work through multi-region indicators downstream. Regions are two rectangles (`R/make-wcvi-region-polygons.R`, saved as a shapefile in `data-raw/wcvi-region-polygons/`) buffered beyond the survey grid extent, so every WCVI cell falls fully inside one rectangle or the other.

* Cells straddling the split line are cut in two by the split; each piece keeps the parent `cell` id but gets its own recalculated `area` (km2) and `region` label, so a split cell contributes one row per region downstream.

```{r}
#| label: wcvi-region-split
cell_id_rast <- empty_raster
names(cell_id_rast) <- "cell"
terra::values(cell_id_rast) <- seq_len(terra::ncell(empty_raster))

wcvi_cells <- depth_df0$cell[depth_df0$survey_abbrev == "SYN WCVI"]
cell_id_rast[!(terra::values(cell_id_rast) %in% wcvi_cells)] <- NA

wcvi_cell_polys <- terra::as.polygons(cell_id_rast, dissolve = FALSE) |>
  sf::st_as_sf() |>
  # Keeps surveys so they can be included as predictors if included in model as factor.
  left_join(terra::as.data.frame(survey_rast, cells = TRUE), by = "cell")

region_polys <- sf::st_read(
  file.path(data_raw_dir, "wcvi-region-polygons", "wcvi-region-polygons.shp"),
  quiet = TRUE
) |>
  sf::st_transform(sf::st_crs(wcvi_cell_polys))

# st_intersection() clips each cell to whichever rectangle(s) it overlaps;
# cells straddling the N/S boundary get one row per region, each with its
# own recalculated area.
wcvi_cell_region_polys <- sf::st_intersection(wcvi_cell_polys, region_polys) |>
  mutate(area = as.numeric(sf::st_area(geometry)) / 1e6) |> # km^2
  filter(area > 1e-6) |> # drop GEOS sliver artifacts along the split line
  group_by(cell, region, survey_abbrev) |>
  summarise(area = sum(area), .groups = "drop")

depth_df0 |>  filter(is.na(depth_m)) |>  glimpse()

# Make prediction grid data.frame (keep this object lean because it is expanded to all years)
prediction_grid <- wcvi_cell_region_polys |>
  sf::st_drop_geometry() |>
  mutate(
    # @Note - centroids must come from the original (unsplit) raster grid:
    # cells straddling the N/S split line are cut into two region polygons,
    # so a centroid taken from the split polygon would be off-center from
    # the true cell
    X = terra::xyFromCell(empty_raster, cell)[, "x"] / 1000,
    Y = terra::xyFromCell(empty_raster, cell)[, "y"] / 1000
  ) |>
  left_join(depth_df0 |> select(cell, depth_m, log_depth), by = "cell") |>
  left_join(temp_df0 |> select(cell, year, bottom_temp), by = "cell", relationship = "many-to-many") |>
  filter(!is.na(bottom_temp), !is.na(log_depth), !is.infinite(log_depth))

# Check prediction grid
wcvi_cell_region_polys |>
  select(cell, region, geometry) |>
  left_join(
    prediction_grid |> filter(year == max(dog$year) + 1),
    by = c("cell", "region")
  ) |>
  ggplot() +
  geom_sf(aes(fill = depth_m, colour = region)) # @CHECK - Why are there missing depths in Trond's dataset? We should probably grab these from Angelica's?
  # geom_sf(aes(fill = bottom_temp, colour = region))

unique(prediction_grid$year)

```

### Save prepared data for species distribution modelling

```{r}
out_filename <- "dogfish.rds" # likely replace with species name from raw data

# Objects saved
# - cleaned survey data
saveRDS(dog, file = file.path(model_input_dir, out_filename))

# - minimize object sizes: rasterized cell polygons (keyed by `cell`/`region`)
# - join downstream for plotting / sf uses
# @TODO: might make sense to make this a standard filename
saveRDS(wcvi_cell_region_polys, file = file.path(model_input_dir, "wcvi-cell-polygons.rds"))

# - prediction grid data.frame
saveRDS(prediction_grid, file = file.path(model_input_dir, "prediction-grid.rds"))
```

* Species survey data with covariate data (bottom temperature and depth) [data.frame]

* Prediction grid (depth + bottom temperature, all forecast years) [data.frame], keyed by `cell`/`region`

* Rasterized cell polygons keyed by `cell`/`region` [sf], for joining predictions back to polygons for plotting