---
title: "1. Data Preparation"
format: html
execute:
  echo: false      # hide code
  warning: false    # hide warnings
  message: false    # hide messages (e.g. package load chatter)
---

```{r}
#| include: false
knitr::opts_chunk$set(results = "hide")
```

```{r}
# Setup
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(terra)
library(sf)

source(here::here("R", "00-settings.R"))

```

### 1. Data preparation

* Using placeholder statistically downscaled dataset for bottom temperature, and matched it to the survey grid (prediction grid). Median bottom temperature in June.

* Combine temperature data with the dogfish survey records

* Visual check: map dogfish observations and temperature patterns by year for the full survey and for one region (WCVI)

```{r}
# Prepare dogfish data, including temperature
dog0 <- dogfish # from sdmTMB package
dog <- dog0 |> mutate(log_depth = log(depth))

# Scoped to the survey's historical year range for now (fitting only needs
# observed years). `temp_dat0` actually holds columns for 2002-2099 across
# climate scenarios (see comment below) - when forecasting is added, widen
# this to the full projection year range (and select a scenario/percentile -
# once more added) here and in `value_cols` below.
historical_years <- min(dog$year):max(dog$year)

# Trond dataset - processed by Sean:
# - extracted data for 2002-2099 for one climate scenario for one month (june) and the 50th percentile value only, that works out to a ~70 mb .tif file
# - interpolated it onto the synoptic grid using terra::extract() given the survey grid centroids
# - 1 climate scenario at the median across the model ensemble
temp_dat0 <- readRDS(file.path(data_raw_dir, "thetao_ensemble_sd+ba_bottom_perc_ssp126_p50_survey_blocks_active.rds"))

grid_vect <- terra::vect(temp_dat0) # Convert temp_dat0 to terra SpatVector to be able to use `terra::extract()`
empty_raster <- terra::rast(grid_vect, resolution = 2000) # Empty array that matches dim of temp_dat0

# Subset temp_dat0 to only historical years (for efficiency)
value_cols <- paste0("value_", historical_years, "_06") # (06 = June)
temp_stack <- terra::rast(lapply(value_cols, \(v) {
  terra::rasterize(grid_vect, empty_raster, field = v)
}))
names(temp_stack) <- historical_years

dog_vect <- terra::vect(dog, geom = c("longitude", "latitude"), crs = "EPSG:4326") |>
  terra::project(terra::crs(temp_stack))

dog$bottom_temp <- terra::extract(
  temp_stack,
  dog_vect,
  method = "bilinear",
  layer = as.character(dog$year)
)$value

glimpse(dog)

dog_sf <- dog |> sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

```{r}
#| label: wcvi-survey-set-bottom-temp

ggplot(dog_sf) +
  geom_sf(data = pacea::bc_coast) +
  geom_sf(aes(colour = bottom_temp)) +
  viridis::scale_colour_viridis(option = "plasma") +
  facet_wrap(~ year) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
  theme(legend.position = "bottom") +
  ggtitle("Survey set bottom temperatures") +
  viridis::scale_colour_viridis(
    option = "plasma",
    name = "Bottom temperature (°C)",
    guide = guide_colourbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.4, "cm"),
      title.vjust = 0.9,
      direction = "horizontal"
    )
  ) +
  gfplot::coord_sf_auto(dog_sf, buffer = 10000)
```

```{r}
# Save covariate data to use when building prediction grids
# - dp before we filter to WCVI, so we can scale up later
# - keep `cell` (terra's integer cell index) alongside X/Y - temp_df and
#   depth_df are built from separate as.data.frame() calls, and joining on
#   the X/Y doubles later silently drops rows where the two don't compare
#   exactly equal. `cell` is an exact integer key since both rasters share
#   the same `empty_raster` template.
# TODO: will need to replace with more complete depth layer;
depth_rast <- terra::rasterize(grid_vect, empty_raster, field = "depth_m")
survey_rast <- terra::rasterize(grid_vect, empty_raster, field = "survey_abbrev")

depth_df <- terra::as.data.frame(c(depth_rast, survey_rast), xy = TRUE, cells = TRUE) |>
  mutate(X = x / 1000, Y = y / 1000, log_depth = log(depth_m)) |>
  select(cell, X, Y, depth_m, log_depth, survey_abbrev)

# survey_abbrev doesn't vary by year, so attach it via a left_join on `cell`
# (not filtering anything out here - just enriching every temp_df row so it
# can be filtered to a region the same simple way as depth_df downstream).
temp_df <- terra::as.data.frame(temp_stack, xy = TRUE, cells = TRUE) |>
  tidyr::pivot_longer(-c(cell, x, y), names_to = "year", values_to = "bottom_temp") |>
  mutate(year = as.integer(year), X = x / 1000, Y = y / 1000) |>
  left_join(depth_df |> select(cell, survey_abbrev), by = "cell") |>
  select(cell, X, Y, year, bottom_temp, survey_abbrev)
```

```{r}
#| label: trond-bottom-temp

# crop for faster plotting of WCVI only data
wcvi_vect <- terra::vect(temp_dat0[temp_dat0$survey_abbrev == "SYN WCVI", ])
wcvi_temp <- terra::crop(temp_stack, wcvi_vect, mask = TRUE)

ggplot() +
  tidyterra::geom_spatraster(data = wcvi_temp) +
  geom_sf(data = pacea::bc_coast) +
  facet_wrap(~lyr) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
  theme(legend.position = "bottom") +
  ggtitle("Oceanographic data: Median annual June bottom temperature") +
  viridis::scale_fill_viridis(
    option = "plasma",
    name = "Bottom temperature (°C)",
    na.value = "transparent",
    guide = guide_colourbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.4, "cm"),
      # title.position = "top",
      title.vjust = 0.9,
      direction = "horizontal"
    )
  ) +
  gfplot::coord_sf_auto(dog_sf, buffer = 10000) # zoom into WCVI
```

### Save prepared data for species distribution modelling

```{r}
out_filename <- "dogfish.rds" # likely replace with species name from raw data

# @TODO: ADD REGION

# Objects saved
# - cleaned survey data
saveRDS(dog, file = file.path(model_input_dir, out_filename))

# - temperature data grid (historical years only right now)
saveRDS(temp_df, file = file.path(model_input_dir, "bottom-temperature-df.rds"))

# - depth data grid
saveRDS(depth_df, file = file.path(model_input_dir, "depth-df.rds"))

```

* Species survey data with covariate data (bottom temperature and depth) [data.frame]

* Temperature for prediction grid [data.frame] (historical only right now; todo: add forecasting)

* Depth for prediction grid [data.frame]