---
title: "2. Model Fitting"
format: html
execute:
  echo: false      # hide code
  warning: false    # hide warnings
  message: false    # hide messages (e.g. package load chatter)
---

```{r}
#| include: false
knitr::opts_chunk$set(results = "hide")
library(dplyr)
library(ggplot2)
library(sdmTMB) # @sdmTMB::project conflicts with terra::project (I use namespaced terra calls in R/01-data-prep.qmd)
library(sf)

source(here::here("R", "00-settings.R"))
source(here::here("R/functions.R"))

dog <- readRDS(file.path(model_input_dir, "dogfish.rds"))
prediction_grid <- readRDS(file.path(model_input_dir, "prediction-grid.rds"))
cell_polygons <- readRDS(file = file.path(model_input_dir, "wcvi-cell-polygons.rds"))

```

### 2. Fit species distribution model using sdmTMB

* Fit species distribution model to model dogfish catch rates as a function of depth and bottom temperature

* Autoregressive models to allow for forecasting

```{r}
dog_mesh <- make_mesh(data = dog, xy_cols = c("X", "Y"), cutoff = 30) # can make smaller later

missing_years <- sdmTMB:::find_missing_time(dog$year) # why not export this?

# Have two types of models so I can make sure code works with one or two linear predictors
fit_tw <- local(
  sdmTMB(
    formula = catch_weight ~ 1 + poly(log_depth, 2) + bottom_temp,
    offset = log(dog$area_swept), # include as smoother later?
    # family = delta_gengamma(type = "poisson-link"), # wouldn't fit/taking forever
    family = tweedie(link = "log"),
    data = dog,
    mesh = dog_mesh,
    spatial = "on",
    spatiotemporal = "ar1",
    time = "year",
    extra_time = missing_years,
    control = sdmTMBcontrol(collapse_spatial_variance = TRUE),
    silent = FALSE
  )
)
fit_tw

# fit_tv <- local(update(fit, time_varying = ~ 1, time_varying_type = "ar1"))
# sanity(fit_tv)

fit_ln <- local(
  sdmTMB(
    formula = catch_weight ~ 1 + poly(log_depth, 2) + bottom_temp,
    offset = log(dog$area_swept), # include as smoother later?
    family = delta_lognormal(type = "poisson-link"), # @QUESTION: Should this always be a poisson link where possible?
    data = dog,
    mesh = dog_mesh,
    spatial = "off", # on collapsed (I'm pretty sure but now I'm second guessing and maybe was just taking forever to run - revisit)
    spatiotemporal = "ar1",
    time = "year",
    extra_time = missing_years,
    control = sdmTMBcontrol(collapse_spatial_variance = TRUE),
    silent = FALSE
  )
)
sanity(fit_ln)

fit_ln_tv <- local(
  update(
    fit_ln, time_varying = ~ 1, time_varying_type = "ar1"
  )
)

fit_dgg <- local(
  update(
    fit_ln, family = delta_gengamma(type = "poisson-link")
  )
)

# I SWEAR THIS MODEL WOULD NOT FIT LAST WEEK WHEN I TRIED?!?!?!?!?!?!?! but now it works great and the dgg is awesome?!
fit_dgg_tv <- local(
  update(
    fit_ln, family = delta_gengamma(type = "poisson-link"),
    time_varying = ~ 1, time_varying_type = "ar1"
  )
)
beepr::beep()

fit_ln_tv
sanity(fit_ln_tv)

AIC(fit_tw, fit_ln, fit_dgg, fit_ln_tv, fit_dgg_tv)

```

```{r}
# TEMPORARY - save two models one vs two linear predictor
fit_tw |> saveRDS(file.path(fit_dir, "dogfish-tweedie.rds"))
fit_ln_tv |> saveRDS(file.path(fit_dir, "dogfish-ln-tv.rds"))
fit_dgg |> saveRDS(file.path(fit_dir, "dogfish-dgg.rds"))

```


### 1b. Todo: model diagnostics

```{r}
# Temporary load models so I don't have to run above model fits to look at diagnostics
# fit <- readRDS(file.path(fit_dir, "dogfish-tweedie.rds"))
fit <- readRDS(file.path(fit_dir, "dogfish-ln-tv.rds"))

# @Note: missing years/alternating years/seesaw - from what I saw in power analysis, could this be a problem here? see note in diagnostics section of @R/02-fit-model.qmd
```

* see forthcoming ICES working group diagnostics output

* Should this also include the seesaw index ACF moving window and max biennial oscillation worth including in this? especially when we consider integrating multiple surveys? Or when we have surveys with biennial sampling? I guess I'm thinking this because of our interest in estimating trends in the indicators and I haven't wrapped my head around waht getting it wrong at this stage would mean for them.

```{r}
# @TODO: make this work for the single linear predictor models as well
# - don't feel like doing this right now, want to get to indicators
dog$resid <- residuals(fit, type = "mle-mvn", model = 2)
# note: NA residuals are when catch_weight = 0 (i.e., binomial part of the model)

ggplot(dog, aes(log_depth, resid)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point()

ggplot(dog, aes(bottom_temp, resid)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point()

ggplot(dog, aes(X, Y, colour = resid)) +
  facet_wrap(~ year) +
  geom_point() +
  coord_fixed() +
  scale_colour_gradient2(na.value="grey") +
  ggtitle("Spatial residuals")

s <- simulate(fit, nsim = 300, type = "mle-mvn")
dharma_residuals(s, fit)

```

```{r}
visreg_delta(fit, xvar = "log_depth", model = 1, scale = "response")
visreg_delta(fit, xvar = "log_depth", model = 2, scale = "response")

visreg_delta(fit, xvar = "bottom_temp", model = 1, scale = "response")
visreg_delta(fit, xvar = "bottom_temp", model = 2, scale = "response")
```

#### Mapping predictions (historical data)
```{r predict}
historical_years <- min(dog$year):max(dog$year)

predictions <- predict(
  fit,
  newdata = prediction_grid |> filter(year %in% historical_years),
  type = "response"
)

pred_poly <- left_join(cell_polygons, predictions)

# Some visual checks -----------------------------------------------------------
# Also useful diagnostic to see if NA values make sense - wrt to where we are truly missing covariate data
ggplot(pred_poly) +
  geom_sf(aes(fill = est), colour = NA) +
  scale_fill_viridis_c(trans = "log10") +
  facet_wrap(~year) +
  ggtitle("Predicted distribution of Pacific Spiny Dogfish (WCVI)")
# ggtitle("Predicted distribution") # TODO add species name


# @TODO: these plots are a lot, improve their layout, maybe just subset some years, not sure
plot_field(pred_poly, fit, "omega_s", "Spatial random field")
plot_field(pred_poly, fit, "epsilon_st", "Spatiotemporal random field")

```




# Forecasting

* Note - this is not necessary for running R/03-get-shift-indicators.qmd

* Maybe we want to look at forecast diagnostics here?

```{r}
# Temporary load models so I don't have to run above model fits to look at diagnostics
# fit <- readRDS(file.path(fit_dir, "dogfish-tweedie.rds"))
# fit <- readRDS(file.path(fit_dir, "dogfish-ln-tv.rds"))
fit <- readRDS(file.path(fit_dir, "dogfish-dgg.rds"))
```


```{r}
out <- project(fit, newdata = prediction_grid, nsim = 50) # values in link space

glimpse(out)

# @hack to deal with returning of each delta model estimate
if (fit$family$delta) out$est <- fit$family[[1]]$linkinv(out$est1) * fit$family[[2]]$linkinv(out$est2)

est_mean <- apply(out$est, 1, mean) # summarize however you'd like
est_se <- apply(out$est, 1, sd)

# visualize:
prediction_grid$est_mean <- est_mean
ggplot(subset(prediction_grid, year > 2021), aes(X, Y, fill = est_mean)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  scale_fill_viridis_c() +
  ggtitle("Projection simulation (mean)")

```