---
title: "3. Historical Shift Indicators"
format: html
execute:
  echo: false     # hide code
  warning: false  # hide warnings
  message: false  # hide messages
---

```{r setup}
#| include: false
knitr::opts_chunk$set(results = "hide", autodep = TRUE)
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(sf)
library(patchwork)

source(here::here("R", "00-settings.R"))
```

```{r load-model-outputs}
#| include: false
#| cache: true
#| cache.extra: !expr file.mtime(c(file.path(model_out_dir, "fits", "dogfish.rds"), file.path(model_out_dir, "predictions", "dogfish.rds"), file.path(model_out_dir, "predictions-sf", "dogfish.rds")))

fit <- readRDS(file.path(model_out_dir, "fits", paste0("dogfish", ".rds")))

# @TODO: Setup so the two dependencies are:
# 1) fit
# 2) prediction grid with regions column
predictions <- readRDS(file.path(model_out_dir, "predictions", paste0("dogfish", ".rds")))

pred_sf0 <- readRDS(file.path(model_out_dir, "predictions-sf", paste0("dogfish", ".rds"))) # @NOTE THIS RIGHT NOW has the combined estimate
# @TODO: might be kind of confusing to have these outputs be different

# filter(pred_sf0, is.na(est_combined))
pred_sf <- pred_sf0 |>  filter(!is.na(bottom_temp) | !is.na(log_depth)) # these are grid cells that did not have covariate data


# Simulated predictions
pred_grid <- predictions$data |> select(X, Y, year, bottom_temp, log_depth, survey_abbrev, area)

# @CHECK: I think we need this for getting area occupied?
set.seed(1)
sims_combined_response <- predict(fit, newdata = pred_grid, type = "response", nsim = 200, model = NA) # sims: matrix, nrow(nd) x 200 — response-scale combined abundance
head(sims_combined_response) #remind myself what this looks like

# @CHECK: But we need this for get_range_edge?
sims_link <- predict(fit, newdata = pred_grid, type = "link", nsim = 200)


# WARNING! I fear that utm zones will bite me later - make sure to set this up safely
```

### 3. Get shift indicators

* This is for 'full region', we can split the WCVI region into two polygons to workthrough multi-region indicators/presentation

* Question: Need to be able to select by season? --> Yes if there is season

```{r}
ggplot(pred_sf) +
    geom_sf(aes(fill = est_combined, colour = est_combined)) +
    viridis::scale_fill_viridis(trans = "log10") +
    viridis::scale_colour_viridis(trans = "log10") +
    facet_wrap(~ year) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 2)) +
    ggtitle("Predicted distribution of Pacific Spiny Dogfish (WCVI)")
```

#### Abundance trend

Trend in total and regional abundance or biomass over time as estimated by the model: describing the size of a population in terms of the relative number or total weight of individuals

* For dogfish we're modelling biomass (not abundance)

```{r abundance}
# @TODO include by region
# @TODO - I guess we will want to have a set of calculations for each index across the full modelled region and then also the subsets?
abundance0 <- get_index(predictions, bias_correct = TRUE, area = predictions$data$area) # @Note: safest if area is included in predictions data (i.e., if area varies)
abundance <- abundance0 |>
  mutate(centred_year = year - mean(year))

trend_lm <- lm(log(est) ~ centred_year, data = abundance)
summary(trend_lm)
# @TODO: get_index() returns a per-year se but we aren't accounting for the AR1 process when we fit the lm. What is the best solution?
# @Question: error propagation: is one strategy to calculate trends for each nsim to propagate uncertainty of annual estimates? (doesn't account for AR1, or does it?)
# @Question: worth using variance weighting in lm() for now?
# @Note: missing years/alternating years/seesaw - from what I saw in power analysis, is this going to be a problem here? see note in diagnostics section of @R/02-fit-model.qmd
# @Question: what should we be presenting as units for abundance trend summary?

abundance_summary <- broom::tidy(trend_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "abundance",
    name = "Abundance Trend",
    estimate,
    se = std.error,
    units = "log(biomass) / year", # @TODO: generalise the units
    p_value = p.value
  )

abundance_summary


# Abundance trend plot

ggplot(data = abundance) +
  aes(x = year, y = est, ymin = lwr, ymax = upr) +
  geom_line() +
  geom_ribbon(alpha = 0.3)

# Thought lognormal might make the index a bit less messy; I guess not.
```

#### Proportional abundance

The percentage of the total abundance/biomass that falls within a given region. Indicates changes in the relative contribution of a region to the overall abundance/biomass.

```{r proportional-abundance}
# For proportion of total abundance --------------------------------------------
# @Question: best way to propagate uncertainty, since this is a ratio of two estimates? using nsim from same seed comparison?
# - should we be getting the ratio from the nsim values?
total_abundance0 <- get_index(predictions, bias_correct = TRUE, area = predictions$data$area)
total_abundance <- total_abundance0 |> select(year, total_est = est)

prop_abundance_df <- left_join(abundance, total_abundance) |>
  mutate(prop_abundance = est / total_est)

# @TODO: use qlogis as temporary solution once actually dealing with ration != 1; may want to consider something else, betareg?
prop_abundance_lm <- lm(prop_abundance ~ centred_year, data = prop_abundance_df)
# prop_abundance_lm <- lm(qlogis(prop_abundance) ~ centred_year, data = prop_abundance_df) # logit transform for now


# @TODO: add uncertainty
prop_abundance_df |>
  ggplot() +
  aes(x = year, y = prop_abundance) +
  geom_line()

prop_abundance_summary <- broom::tidy(trend_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "proportional-abundance",
    name = "Proportional Abundance Trend",
    estimate,
    se = std.error,
    units = "proportion of total abundance / year", # @TODO: convert to percent?
    p_value = p.value
  )

```

#### Centre of gravity

Abundance-weighted mean latitude and longitude of a species' distribution: describing the location of the centre of the distribution and reflecting shifts in this location over time
```{r get-cog}
#| output: false
#| cache: true

tictoc::tic("get_cog, bias_correct = TRUE")
cog <- get_cog(predictions, format = "wide", bias_correct = TRUE)
tictoc::toc()
# get_cog, bias_correct = TRUE: 74.494 sec elapsed
tictoc::tic("get_cog, bias_correct = FALSE")
cog_no_correct <- get_cog(predictions, format = "wide", bias_correct = FALSE)
tictoc::toc()
# get_cog, bias_correct = FALSE: 6.079 sec elapsed
cog

ggplot(cog, aes(est_x, est_y, colour = year)) +
  geom_point() +
  geom_linerange(aes(xmin = lwr_x, xmax = upr_x)) +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y))

ew <- ggplot(cog, aes(year, est_x)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_x, ymax = upr_x))

ns <- ggplot(cog, aes(year, est_y)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y))

# ns | ew

fit_x <- lm(est_x ~ year, data = cog, weights = 1 / se_x^2) # inverse variance weighting for now @Question - even worth it?
fit_y <- lm(est_y ~ year, data = cog, weights = 1 / se_y^2)

cog_summary <- bind_rows(
  broom::tidy(fit_x) |> filter(term == "year") |> mutate(name = "Centre Shift (E/W)", indicator = 'cog-ew'),
  broom::tidy(fit_y) |> filter(term == "year") |> mutate(name = "Centre Shift (N/S)", indicator = 'cog-ns')
) |>
  transmute(
    indicator,
    name,
    estimate,
    se = std.error,
    units = "km / year",
    p_value = p.value,
  )

cog_summary
```
```{r}
#| fig-width: 7
#| fig-height: 8

cog_sf <- XY_to_sf(cog, coords = c("est_x", "est_y"), crs_to = st_crs(pred_sf)) |>
  arrange(year)

coords <- sf::st_coordinates(cog_sf)
n <- nrow(coords)

cog_segments_sf <- sf::st_sf(
  year = cog_sf$year[-n],
  year_end = cog_sf$year[-1],
  geometry = sf::st_sfc(
    purrr::map(seq_len(n - 1), \(i) sf::st_linestring(coords[i:(i + 1), , drop = FALSE])),
    crs = sf::st_crs(cog_sf)
  )
)

ggplot() +
  geom_sf(data = pacea::bc_coast, fill = "grey90", linewidth = 0.2) +
  geom_sf(data = pred_sf |> filter(year == max(pred_sf$year)), aes(fill = est_combined), alpha = 0.4, colour = NA) +
  scale_fill_viridis_c(trans = "log10") +
  # geom_sf(data = cog_segments_sf, aes(colour = year), linewidth = 1) +
  geom_sf(data = cog_segments_sf, aes(colour = year), linewidth = 0.8,
        arrow = grid::arrow(length = unit(0.18, "cm"), type = "closed")) +
  geom_sf(data = cog_sf |> slice_max(year), aes(colour = year), size = 3) +
  scale_colour_viridis_c(option = "magma", end = 0.7) +
  labs(colour = "Year", title = "Centre of gravity trajectory, Pacific Spiny Dogfish (WCVI)") +
  gfplot::coord_sf_auto(pred_sf)

# ggplot() +
#   geom_sf(data = pred_sf, aes(fill = est_combined, colour = est_combined))+
#   viridis::scale_fill_viridis(trans = "log10") +
#   viridis::scale_colour_viridis(trans = "log10") +
#   geom_sf(data = cog_segments_sf, colour = "black", alpha = 0.8) +
#   geom_sf(data = cog_current, colour = "black") +
#   scale_x_continuous(breaks = scales::pretty_breaks(n = 2)) +
#   scale_y_continuous(breaks = scales::pretty_breaks(n = 2)) +
#   facet_wrap(~ year, ncol = 4) +
#   theme(legend.position = "inside",
#         legend.position.inside = c(0.85, 0.1))

```

#### Area occupied

Total area containing 90\% of a species' abundance/biomass: describing how spread out a population is and indicating expansion or contraction in regional occupancy through time

Three ways to display it on the shiny app plot:

-[x] Area used (%) (by regions)

-[x] Area used (km2)

-[x] Area Efficiency

```{r}
#| output: false
avg_density_eao <- get_eao(predictions, bias_correct = TRUE, area = predictions$data$area) # uses calculation from Thorson et al. 2016 (Density-dependent changes in effective area occupied for sea-bottom-associated marine fishes)

avg_density_eao

ggplot(data = avg_density_eao) +
  aes(x = year, y = est, ymin = lwr, ymax = upr) +
  geom_line() +
  geom_ribbon(alpha = 0.3) +
  ggtitle("Area required to contain population given average population density (sdmTMB built-in)")

```
```{r eao-sims}
#| cache: true

# @TODO add region column to upstream data (in dogfish example will be survey_abbrev for now, or split polygon names)
# DSAF area occupied = "what area holds X%  of the population"
# X = 50, 75, 90, 95 quantiles of population density

# @QUESTION: should combined estimates be included in the `prediction` object?
# @TODO: make sure all units are correct in data prep stage

# modified from: https://github.com/kifre1/Hali_Shift_withNF/blob/665ebe5246f40c65ef8982eb75f0a22bf62097eb/Data/Data_SHinyApp_Proof_of_Concept/Prep_timeseries_and_scaled_slope_data_POC.R#L70
# - without uncertainty

# @TODO: delete? if we use nsim method below?
# cumulative_density_df <-
#   predictions$data |>
#   mutate(
#     abundance = exp(est1 + est2), # @Note - WARNING - this only works if the links are the same (i.e., type = poisson-link)
#     density = abundance / area
#   ) |>
#   # group_by(region, year) # @TODO add region
#   group_by(year) |>
#   # arrange(region, year, desc(density)) |> # @TODO add region
#   arrange(year, desc(density)) |>
#   mutate(
#     total_abundance = sum(abundance),
#     total_area = sum(area),
#     cumulative_abundance = cumsum(abundance),
#     percent_abundance = cumulative_abundance / total_abundance * 100,
#     cumulative_area = cumsum(area)
# ) |>
#   mutate(id = row_number()) # for plotting accumulation curves
#
# ggplot(data = cumulative_density_df, aes(x = id, y = cumulative_abundance)) +
#   geom_point(aes(colour = year))
#
# ggplot(data = cumulative_density_df, aes(x = id, y = cumulative_area)) +
#   geom_point(aes(colour = year))
#
# thresholds <- c(50, 75, 90, 95)
#
# quantile_eao <- purrr::map_dfr(thresholds, \(threshold) {
#   cumulative_density_df |>
#     filter(percent_abundance >= threshold) |>
#     slice(1) |> # take first value after crossing the threshold
#     mutate(
#       threshold = threshold,
#       area_threshold = cumulative_area,
#       percent_area_used = area_threshold / total_area * 100
#     )
#   }
# )
#
# ggplot(data = quantile_eao) +
#   aes(x = year, y = area_threshold, colour = threshold, groups = factor(threshold)) +
#   geom_line() +
#   scale_colour_viridis_c(option = "plasma", end = 0.7, direction = -1) +
#   labs(colour = "% of population") +
#   ggtitle("Minimum area needed to contain X% of population")

# Method to get uncertainty? ---------------------------------------------------
# @Question: use predict with nsim > 0 to get with some uncertainty?
# @Question: I am mxing up whether to use link or response values - double check
thresholds <- c(50, 75, 90, 95)

# @TODO make sure that varying areas are not going to be problematic
# @TODO if keeping; pull out and document
get_area_thresholds <- function(df, thresholds) {
  df <- df |>
    # predict(..., offset = 0) returns density (biomass / 1 km^2);
    # Rank by density, not abundance, so a cell's rank doesn't depend on its area
    # - split cells carry unequal, non-4-km^2 areas
    mutate(abundance = density * area) |>
    group_by(year) |> # @TODO add region
    arrange(year, desc(density)) |> # @TODO add region
    mutate(
      total_abundance = sum(abundance),
      total_area = sum(area),
      cumulative_abundance = cumsum(abundance),
      percent_abundance = cumulative_abundance / total_abundance * 100,
      cumulative_area = cumsum(area)
    ) |>
    ungroup()

  purrr::map_dfr(thresholds, \(threshold) {
    df |>
      group_by(year) |>
      filter(percent_abundance >= threshold) |> # @Question: ok? get threshold area just before percent_abundance reaches quantile, or should we get area after crossing before hitting the population quantile?
      slice(1) |>
      ungroup() |>
      mutate(threshold = threshold,
             area_threshold = cumulative_area,
             percent_area_used = area_threshold / total_area * 100,
             area_efficiency = area_threshold / total_abundance)
  }) |>
    select(year, threshold, area_threshold, percent_area_used, area_efficiency)
}

sim_results <- purrr::map_dfr(seq_len(ncol(sims_combined_response)), \(i) {
  pred_grid |>
    mutate(density = sims_combined_response[, i]) |>
    get_area_thresholds(thresholds) |>
    mutate(sim = i)
})

quantile_eao_df <- sim_results |>
  tidyr::pivot_longer(
    cols = c("area_threshold", "percent_area_used", "area_efficiency"),
    names_to = "metric",
    values_to = "value"
    ) |>
  group_by(year, threshold, metric) |>
  summarise(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    lwr = quantile(value, 0.025),
    upr = quantile(value, 0.975),
    .groups = "drop"
  )
```
```{r plot-area-occupied}
metric_labels <- c(
  area_threshold = "Total area occupied",
  percent_area_used = "Percent of area used",
  area_efficiency = "Area efficiency"
)

plot_area_metric <- function(df, metric_name, title, thresholds = NULL) {
  df <- df |> filter(metric == metric_name)
  if (!is.null(thresholds)) df <- df |> filter(threshold %in% thresholds)

  df |>
    ggplot() +
  aes(x = year, y = median, ymin = lwr, ymax = upr,
    colour = threshold, fill = threshold, groups = factor(threshold)) +
  geom_line() +
  geom_ribbon(alpha = 0.2, colour = NA) +
  scale_colour_viridis_c(option = "plasma", end = 0.7, direction = -1) +
  scale_fill_viridis_c(option = "plasma", end = 0.7, direction = -1) +
    labs(colour = "% of population", fill = "% of population", y = NULL, title = title) +
  scale_x_continuous(expand = FALSE) +
  facet_wrap(~ threshold)
}

area_plots <- purrr::imap(metric_labels, \(title, metric_name) {
  plot_area_metric(quantile_eao_df, metric_name, title, threshold = c(90)) # filter threshold here
})

patchwork::wrap_plots(area_plots, ncol = 1) +
  patchwork::plot_layout(guides = "collect")

eao <- quantile_eao_df

eao_90_lm <- lm(median ~ centred_year,
  # weights = 1 / sd^2, # drop if using multiple nsim strategy
  data = eao |> filter(threshold == 90, metric == "area_threshold") |> mutate(centred_year = year - mean(year)))
summary(eao_90_lm)
# @FIXME: current trend method limitations:
#  1) estimate uncertainty (coarsely can use variance weighted as currently done)
#  2) doesn't account for AR1 correlation
# @Question: could fit trend to each nsim and summarise trend estimates from these?
eao_summary <- broom::tidy(eao_90_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "eao-90",
    name = "Area Occupied",
    estimate,
    se = std.error,
    units = "km^2^ / year", # Not sure if this superscript is going to trip us up...
    p_value = p.value
  )

```

#### Range edge

* Change in the position of the southernmost (trailing, 5% quantile) range edge and northernmost (leading, 95% quantile) range edge of modeled species' distribution

* Trailing edge (5% quantile): the latitude beyond which (southward) no more than
  5% of the population's (density-weighted) abundance extends.

* Leading edge (95% quantile): the latitude beyond which (northward) no more than
  5% of the population's (density-weighted) abundance extends.

* What is this really meaning when we have only a small part of the distribution??


```{r range-edges}
head(sims_link)

# @Question: Are we doing E/W edge? This looks like it is in the shiny app indicator wheel.

# @Question: should this be density-weighted? For this example, and our Pacific synoptic surveys this isn't a problem, because all of the grid cells are the same size, but I don't know if this will hold true, or if for example we were to consider the Pacific HBLL, those grid cells overlap with land and so have different overwater areas.
edges <- get_range_edge(sims_link, axis = pred_grid$Y, quantiles = c(0.05, 0.95))

ggplot(edges, aes(year, y = est, ymin = lwr, ymax = upr, group = quantile)) +
  geom_ribbon(fill = "grey90") +
  geom_line()

# @Question: do we need to scale year? everything is a rate in years so probably not??
# @Question: should years be centred to final observed year if forecasting?
# @Question: should year be centred upstream - is there any chance of this borking later?
# @Question: should we be filtering/calculating without dplyr?
# @Question: should we assume the edges to act independently?
# @FIXME: current trend method limitations:
#  1) estimate uncertainty (coarsely can use variance weighted as currently done)
#  2) doesn't account for AR1 correlation
# @Question: could fit trend to each nsim and summarise trend estimates from these?

north_edge_lm <- lm(est ~ centred_year, data = edges |> filter(quantile == 0.95) |> mutate(centred_year = year - mean(year)))
summary(north_edge_lm)

south_edge_lm <- lm(est ~ centred_year, data = edges |> filter(quantile == 0.05) |> mutate(centred_year = year - mean(year)))
summary(south_edge_lm)

north_edge_summary <- broom::tidy(north_edge_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "edge-north",
    name = "Edge Shift (N)",
    estimate,
    se = std.error,
    units = "km / year",
    p_value = p.value
  )

south_edge_summary <- broom::tidy(south_edge_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "edge-south",
    name = "Edge Shift (S)",
    estimate,
    se = std.error,
    units = "km / year",
    p_value = p.value
  )

edge_summary <- bind_rows(north_edge_summary, south_edge_summary)

```


#### Average depth

* Density-weighted average depth (50th percentile). Indicates changes in depth niche over time

```{r mean-depth}
# @Question - should we be using nsim method to get quantiles?

mean_log_depth <- get_weighted_average(predictions, area = predictions$data$area, vector = predictions$data$log_depth)
mean_depth <- get_weighted_average(predictions, area = predictions$data$area, vector = exp(predictions$data$log_depth))

ggplot(data = mean_depth) +
  aes(x = year, y = est, ymin = lwr, ymax = upr) +
  geom_line() +
  geom_ribbon(alpha = 0.3) +
  scale_y_reverse() +
  labs(x = "Year", y = "Mean depth")

depth_lm <- lm(est ~ centred_year, data = mean_depth |> mutate(centred_year = year - mean(year)))
summary(depth_lm)
# @FIXME: same trend fit issue: ignores AR1 cross-year correlation and uncertainty of estimates

```

```{r}
#| output: false

# @Question - do we want median raw depth? or mean + se for log-transformed depth from get_weighted_average?
# @Question - double checking that this should be done for depth (not log transformed)
ggplot() +
  geom_pointrange(data = mean_log_depth, aes(x = year, y = est, ymin = lwr, ymax = upr), colour = "red") +
  geom_pointrange(data = mean_depth, aes(x = year + 0.2, y = log(est), ymin = log(lwr), ymax = log(upr)))

# some diagnostics on the depth trend
plot(depth_lm)

depth_diag <- broom::augment(depth_lm)

ggplot(depth_diag, aes(.fitted, .std.resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted", y = "Standardized residual")

ggplot(depth_diag, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line()

depth_summary <- broom::tidy(depth_lm) |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "depth",
    name = "Depth Trend",
    estimate,
    se = std.error,
    units = "m / year",
    p_value = p.value
  )

depth_summary

```

#### Outputs for shiny app indicator wheel

Note - these are results from simple linear models - the response has not been scaled with `scale()`.

```{r, results=TRUE}

# @Question: should abundance/biomass trend be reported as % change / year instead of the raw log-scale slope?
bind_rows(eao_summary, abundance_summary, cog_summary, edge_summary, prop_abundance_summary, depth_summary) |>
  mutate(
    signif = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      .default = ""
    ),
    across(c(estimate, se), \(x) round(x, 2)),
    p_value = scales::pvalue(p_value)
  ) |>
  knitr::kable(
    col.names = c("Indicator", "Name", "Estimate", "SE", "Units", "p-value", "Sig."), # @TODO: will need to add region column
    align = "llrrlrc"
  )

```
