---
title: "3. Shift Indicators"
format:
  pdf:
    code-overflow: wrap
    include-in-header:
      text: |
        \usepackage{fvextra}
        \fvset{breaklines,breakanywhere}
execute:
  echo: false     # hide code
  warning: false  # hide warnings
  message: false  # hide messages
---

```{r setup}
#| include: false
knitr::opts_chunk$set(results = "hide", autodep = TRUE)
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(sf)
library(patchwork)
library(broom)
library(kableExtra)

source(here::here("R", "00-settings.R"))
source(here::here("R", "functions.R"))
```

```{r load-model-outputs}
fit <- readRDS(file.path(fit_dir, paste0("dogfish-dgg-rw", ".rds")))
# fit <- readRDS(file.path(fit_dir, paste0("dogfish-dgg", ".rds")))
# fit <- readRDS(file.path(fit_dir, paste0("dogfish-tweedie", ".rds")))
prediction_grid0 <- readRDS(file.path(model_input_dir, "prediction-grid.rds")) |>
  filter(year <= 2040)
cell_polygons <- readRDS(file.path(model_input_dir, "wcvi-cell-polygons.rds"))

# Multiple regions not required to work
region_polys_path <- file.path(data_raw_dir, "wcvi-region-polygons", "wcvi-region-polygons.shp")
region_polys <- if (file.exists(region_polys_path)) {
  sf::st_read(region_polys_path, quiet = TRUE)
} else {
  cli::cli_warn("Region polygon shapefile not found at {.file {region_polys_path}}; skipping region overlays.")
  NULL
}

```

Three objects feed `02-fit-model.qmd` and `03-get-shift-indicators.qmd`. All are
built in `01-data-prep.qmd` and cached to disk so downstream scripts can
`readRDS()`/`st_read()` them.

**1. `fit` - sdmTMB model output**

* This is something that should be able to be recalculated from the data
inputs because reloading `sdmTMB` objects built with older versions of `sdmTMB`
break

* Note - the script below assumes that the model was not fit with projections and
that instead, projections are done here using `project()`. This could be adjusted.


**2. `prediction_grid` — plain data.frame, no geometry**

```{r}
#| results: markup
head(prediction_grid0)
```

One row per `cell` x `region` x `year`. This is the `newdata` passed to
`predict()`/`simulate()`/`project()`, so it carries covariates.
All covariates that are in the model should be in this grid.
Rows with missing/infinite depth or missing covariate (temperature) are dropped
upstream, so `NA` covariates shouldn't appear here.

*Minimum required* — every `newdata` grid needs these regardless of model
(key, coordinates, area weighting, and the time index):

| column   | type    | notes                                            |
|----------|---------|--------------------------------------------------|
| `cell`   | numeric | grid cell id (raster cell number)                |
| `region` | chr     | `"WCVI N"` / `"WCVI S"`                          |
| `X`, `Y` | numeric | cell centroid, km, UTM 9N                        |
| `area`   | numeric | cell area, km² (recalculated for split cells)    |
| `year`   | int     | survey year or forecast year                     |

*Model-specific covariates* — depend on the fitted formula, included here
because this model uses depth and bottom temperature:

| column          | type    | notes                                                 |
|-----------------|---------|-------------------------------------------------------|
| `survey_abbrev` | factor  | survey id for each grid cell (not actually used here) |
| `depth_m`       | numeric | bottom depth, m                                       |
| `log_depth`     | numeric | `log(depth_m)`                                        |
| `bottom_temp`   | numeric | bottom temperature covariate, °C                      |

For example in the dogfish demo:


**3. `cell_polygons` — sf object, one row per cell x region**
```{r}
#| results: markup
cell_polygons |> head()
```

* I think this is useful for plotting because we can use `sf` ggplot2 methods

* Same key/covariate columns as `prediction_grid` minus the per-year fields
(`year`, `bottom_temp`), plus the polygon `geometry`. This is what you
`left_join()` predictions back onto for mapping (see `pred_poly` in
`02-fit-model.qmd`).

| column          | type       | notes                                           |
|-----------------|------------|-------------------------------------------------|
| `cell`          | numeric    | grid cell id                                    |
| `region`        | chr        | region names                                    |
| `survey_abbrev` | factor     | survey names                                    |
| `area`          | numeric    | cell area, km²                                  |
| `geometry`      | POLYGON sf | cell footprint, CRS UTM, but could be different |

3,520 rows.

**4. `region_polys` — sf object, one row per region**

```{r}
#| results: markup
region_polys
```

The two N/S polygons from a shapefile used to split the WCVI grid (built by
`R/make-wcvi-region-polygons.R`).
Used for overlay in maps (`geom_sf`) and to derive region boundary lines,
not for per-cell joins.
Was used in `R/01-data-prep.R` to split the 'all' grid into regions.

| column     | type       | notes                                |
|------------|------------|--------------------------------------|
| `region`   | chr        | region name                          |
| `geometry` | POLYGON sf | buffered bounding rectangle, UTM 9N  |


### Data preparation
```{r}
# Get line segments for region borders to declutter plots downstream
region_borders <- get_region_borders(region_polys)
region_layer <- if (!is.null(region_polys)) geom_sf(data = region_polys, fill = NA, linewidth = 0.5)

last_historical_year <- max(fit$data$year) # use for centring so no ambiguity in lm trend intercept
```

```{r questions}
#| echo: false
# Predictions conditioned on best estimate of random effects
# @Question - what words/phrasing do I use to describe the predictions using
# nsim = 0 vs nsim >0? expected values given best MLEs of both fixed and random
# effects vs simulated prediction draws sampled from joint precision matrix?
# @Question - Should we just be using OSP draws of random effects for all of
# the shift indicators? The problem right now is we need to use a one sample
# posterior draw of random effects NOT the joint precision matrix to avoid biased
# estimate, correct? And the mechanics of doing this would be that we need to
# use `simulate_new()`(?) to get our draws and manually take our own OSP? Can we
# do this with project and supply the OSP?

# See: https://github.com/sdmTMB/sdmTMB/discussions/431#discussioncomment-12717246
# See: https://github.com/sdmTMB/sdmTMB/discussions/431

# We are not currently simulating from the spatial random effects (i.e., not
# accounting for uncertainty in the sp rf) - worth asking Andrew - may also be
# in the fish and fisheries paper (this is the first `0` in the sim_re)

```

Generate simulated draws; right now separately done using `simulate()` for the
historical data and `project()` for forecasts. No observation error included.

```{r sim-draws}
#| cache: true
#| cache.extra: !expr tools::md5sum(c(file.path(fit_dir, "dogfish-dgg.rds"), file.path(model_input_dir, "prediction-grid.rds")))
# @Note: only safe to log if `observation_error = FALSE` because `simulate() returns the expected density plogis(η₁)·exp(η₂), which is strictly positive (plogis output cannot be 0). Zeros only come from the presence/absence draw when `observation_error = TRUE`.

set.seed(2026)

prediction_grid <- prediction_grid0 |> filter(year <= 2035)

projections <- project(
  fit,
  newdata = prediction_grid,
  nsim = 400,
  sample_fe = FALSE,
  sample_historical_re = TRUE,
  sample_future_re = TRUE,
  offset = rep(0, nrow(prediction_grid))
)

# projections_fe <- project(
#   fit,
#   newdata = prediction_grid,
#   nsim = 400,
#   sample_fe = FALSE,
#   sample_historical_re = TRUE,
#   sample_future_re = TRUE,
#   future_re = "fix",
#   offset = rep(0, nrow(prediction_grid))
# )

meep()

# projections_re <- project(
#   fit,
#   newdata = prediction_grid |> filter(year <= 2035),
#   nsim = 200, # change to >100 for demo
#   sample_parameters = TRUE, # if FALSE: theta at mode, b at mode → lp = lpb exactly, replicated. Zero stochasticity = single line.
#   sample_historical_re = TRUE,
#   sample_future_re = TRUE,
#   sims_var = "eta_i",
#   simulate_re = c(
#       spatial = FALSE,
#       spatiotemporal = TRUE,
#       spatial_varying = FALSE,
#       iid = FALSE,
#       time_varying = TRUE, # dgg model only doesn't have any
#       smoothers = FALSE
#     )
# )
meep()

projection_log_density <- projections$est1 + projections$est2
# projection_fe_log_density <- projections_fe$est1 + projections_fe$est2

sims_re <- make_sims_df(
    prediction_grid,
    projection_log_density,
    last_historical_year
  ) |>
    mutate(type = "re")

  sims_fe <- make_sims_df(
    prediction_grid,
    projection_fe_log_density,
    last_historical_year
  ) |>
    mutate(type = "fe")

sims <- sims_re
# sims <- sims_re

saveRDS(sims_re, file.path(model_out_dir, "sims_re.rds"))
saveRDS(sims_fe, file.path(model_out_dir, "sims_fe.rds"))

```

Visualise the species distributions for the historical and full series.
```{r pred-map-historical}
#| fig-width: 8
#| fig-height: 13

# Predicted values plot
# @TODO - add bathymetric layer
sims |>
  filter(time_period == "historical") |>
  group_by(time_period, year, cell, region) |>
  filter(sim == 3) |>
  filter(year %in% seq(2004, 2022,  by = 4)) |>
  summarise(log_density = median(log_density)) |>
  left_join(cell_polygons, y = _) |>
  tidyr::drop_na(log_density) |>
ggplot() +
  geom_sf(aes(fill = exp(log_density), colour = exp(log_density))) +
  scale_fill_viridis_c(option = "plasma") +
  scale_colour_viridis_c(option = "plasma") +
  # scale_fill_viridis_c(trans = "log10", option = "plasma") +
  # scale_colour_viridis_c(trans = "log10", option = "plasma") +
  geom_sf(data = region_polys, fill = NA, linewidth = 0.5) +
  facet_wrap(~ year, ncol = 3) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

```{r pred-map-full-series}
#| fig-width: 9
#| fig-height: 13

sims |>
  # filter(time_period == "forecast") |>
  # filter(year %in% seq(last_historical_year + 1, 2060, by = 2)) |>
  filter(year %in% seq(2005, 2060,  by = 4)) |>
  # filter(sim == 3) |>
  group_by(time_period, year, cell, region) |>
  summarise(log_density = median(log_density)) |>
  left_join(cell_polygons, y = _) |>
  tidyr::drop_na(log_density) |>
ggplot() +
  geom_sf(aes(fill = exp(log_density), colour = exp(log_density))) +
  scale_fill_viridis_c(option = "plasma") +
  scale_colour_viridis_c(option = "plasma") +
  # scale_fill_viridis_c(trans = "log10", option = "plasma") +
  # scale_colour_viridis_c(trans = "log10", option = "plasma") +
  geom_sf(data = region_polys, fill = NA, linewidth = 0.5) +
  facet_wrap(~ year, ncol = 4) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

```{r pred-cell-sf}
# For plotting background - make predictions into sf object
pred_cell_sf <- sims |>
  group_by(time_period, year, cell, region) |>
  summarise(density_km2 = exp(median(log_density)), .groups = "drop") |>
  left_join(cell_polygons, y = _) |>
  tidyr::drop_na(density_km2)
```

### 3. Get shift indicators

Below returns shift indicators by:

* time period (historical vs forecast)
* region (subregions and full domain ["all"])

This could be extended to included additional groupings - e.g., season. We could
define a grouping vector above in the data prep.

#### Abundance trend

Trend in total and regional abundance or biomass over time as estimated by the
model: describing the size of a population in terms of the relative number or
total weight of individuals

* For dogfish we're modelling biomass (not abundance)

```{r abundance}
# @TODO: for lm trends, we aren't accounting for the AR1 process or propagating
# uncertainty - or are we now doing this that we've moved to using nsims > 0?
# @Question: what should we be presenting as units for abundance trend summary dataframe?
# @Question: should we be summarising by mean to match E(X) with bias correction?
# Or median? Or doesn't matter since we're only looking at trends?

# for my reference
# get_index_sims does:
  # est_function = stats::median()
  # area_function = function(x, area) x + log(area)
  # agg_function = function(x) sum(exp(x)) - because values must come logged

# Calculate annual abundance for each simulated draw
a_by_sim <- sims |>
  group_by(time_period, region, year, sim) |>
  summarise(total_abundance = sum(exp(log_abundance)), .groups = "drop")

a_by_region <- a_by_sim |>
  summarise_sim_ci(cols = total_abundance, by = c(time_period, region, year))

abund_lm_by_sim <- a_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  summarise_sim_trend(value = log(total_abundance), by = c(time_period, region))

abundance_summary <- abund_lm_by_sim |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "abundance",
    name = "Abundance Trend",
    region = region,
    time_period = time_period,
    estimate,
    se = se,
    lwr = lwr,
    upr = upr,
    prob_direction,
    units = "log(biomass) / year" # @TODO: generalise the units
  )

a_by_region |>
  filter(region != "all") |>
ggplot() +
  aes(x = year, y = est, ymin = lwr, ymax = upr, colour = region, fill = region) +
  geom_line() +
  geom_ribbon(colour = NA, alpha = 0.3) +
  geom_vline(xintercept = last_historical_year, linetype = "dashed", colour = "grey30") +
  coord_cartesian(expand = FALSE)

a_by_sim |>
  filter(region == "all") |>
ggplot() +
  aes(x = year, y = log(total_abundance), colour = region, fill = region, group = interaction(sim, region)) +
geom_ribbon(data = a_by_region |> filter(region == "all"),
  aes(y = log(est), ymin = log(lwr), ymax = log(upr), group = NULL),
  fill = blues[[1]], colour = NA, alpha = 0.7) +
  geom_line(colour = blues[[4]], alpha = 0.3) +
  geom_line(data = a_by_region |> filter(region == "all"),
    aes(y = log(est), group = NULL),
    colour = blues[[5]], linewidth = 1) +
  geom_vline(xintercept = last_historical_year, linetype = "dashed", colour = "grey30") +
  coord_cartesian(expand = FALSE, clip = "on") +
  facet_wrap(~ region, ncol = 1) +
  ggtitle(unique(sims$type))


```

#### Proportional abundance

The percentage of the total abundance/biomass that falls within a given region.
Indicates changes in the relative contribution of a region to the overall abundance/biomass.

```{r proportional-abundance}
# @Question: best way to propagate uncertainty, since this is a ratio of two estimates? using nsim from same seed comparison does this?
# - should we be getting the ratio from the nsim values?
totals_by_sim <- a_by_sim |>
  filter(region == "all") |>
  select(time_period, year, sim, total_abundance_all = total_abundance)

p_abund_by_sim <- a_by_sim |>
  filter(region != "all") |>
  left_join(totals_by_sim, by = c("time_period", "year", "sim")) |>
  mutate(prop = total_abundance / total_abundance_all)

p_abund <- p_abund_by_sim |>
  summarise_sim_ci(cols = prop, by = c(time_period, region, year))

p_abund_lm <- p_abund_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  summarise_sim_trend(value = qlogis(prop), by = c(time_period, region))

prop_abundance_summary <- p_abund_lm |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "prop_abundance",
    name = "Proportional Abundance Trend",
    region = region,
    time_period = time_period,
    estimate,
    se,
    lwr,
    upr,
    prob_direction,
    units = "" # @Question - halp, thinking in log-odds changes; additional option could be present as percent over time periods?
  )

# @Question - symmetry is good sign because change in proportion should be zero-sum; does that however have implications for how we model the trend?
# - should we modellng historical and forecast trend all in one model?
p_abund_lm |> arrange(term)

ggplot(data = p_abund) +
  aes(x = year, y = est, ymin = lwr, ymax = upr, colour = region, fill = region, group = region) +
  geom_line() +
  geom_ribbon(alpha = 0.3, colour = NA) +
  ggthemes::scale_colour_ptol() +
  ggthemes::scale_fill_ptol() +
  coord_cartesian(expand = FALSE) +
  geom_vline(xintercept = last_historical_year, linetype = "dashed", colour = "grey50") #+
  # facet_wrap(time_period ~ ., ncol = 1, scales = "free_x")

```

#### Centre of gravity

Abundance-weighted mean latitude and longitude of a species' distribution: describing the location of the centre of the distribution and reflecting shifts in this location over time

For reference from Thorson, Pinsky, & Ward 2016:
$$
\text{Latitude}_{\text{COG}} = \frac{\sum_{i=1}^{n_s} \hat{d}(s_i) \cdot \text{Latitude}(s_i)}{\sum_{i=1}^{n_s} \hat{d}(s_i)} \qquad (\text{eqn } 2)
$$

where:
* $n_s$ is the number of locations (e.g., grid cells) used to summarize a spatial domain,
* $\hat{d}(s_i)$ is the predicted density at location $s_i$, and
* $\text{Latitude}(s_i)$ is the latitude for location $s_i$.

BUT - prediction grid could have unequal cell areas, so we weight by abundance $= \hat{d}(s_i) \times A_i$ (predicted density $\times$ cell area) :
$$
\text{X}_{\text{COG}} = \frac{\sum_{i=1}^{n_s} \hat{d}(s_i) \times A_i \times \text{X}(s_i)}{\sum_{i=1}^{n_s} \hat{d}(s_i) \times A_i} \qquad \text{Y}_{\text{COG}} = \frac{\sum_{i=1}^{n_s} \hat{d}(s_i) \times A_i \times \text{Y}(s_i)}{\sum_{i=1}^{n_s} \hat{d}(s_i) \times A_i}
$$

* $\text{X}(s_i)$ is the UTM Easting (longitude) for location $s_i$.
* $\text{Y}(s_i)$ is the UTM Northing (latitude) for location $s_i$.
* $\text{A}(s_i)$ is the Area for location $s_i$.
* $\text{abundance} = \hat{d}(s_i) \times A_i$ is the Area for location $s_i$.


```{r get-cog}
cog_by_sim <- sims |>
  group_by(time_period, region, year, sim) |>
  summarise(
    x = weighted.mean(X, w = exp(log_abundance)), # same as: x = sum(exp(log_abundance) * X) / sum(exp(log_abundance)),
    y = weighted.mean(Y, w = exp(log_abundance)),
    .groups = "drop"
  )

cog <- cog_by_sim |>
  summarise_sim_ci(c(x, y), c(time_period, region, year))

cog_lm <- cog_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  tidyr::pivot_longer(c(x, y), names_to = "coord", values_to = "cog") |>
  summarise_sim_trend(value = cog, by = c(time_period, region, coord))

cog_summary <- cog_lm |>
  filter(term == "centred_year") |>
  transmute(
    indicator = if_else(coord == "x", "cog-ew", "cog-ns"),
    name      = if_else(coord == "x", "Centre Shift (E/W)", "Centre Shift (N/S)"),
    time_period,
    region,
    estimate,
    lwr,
    upr,
    se,
    prob_direction,
    units     = "km / year"
  )

```
```{r}
#| fig-width: 7
#| fig-height: 8
make_cog_segments <- function(points_sf) {
  points_sf <- dplyr::arrange(points_sf, year)
  coords <- sf::st_coordinates(points_sf)
  n <- nrow(coords)
  if (n < 2) return(NULL)  # a single-year group has no segment to draw
  sf::st_sf(
    region      = points_sf$region[-n],
    time_period = points_sf$time_period[-n],
    year        = points_sf$year[-n],
    year_end    = points_sf$year[-1],
    geometry = sf::st_sfc(
      purrr::map(seq_len(n - 1), \(i) sf::st_linestring(coords[i:(i + 1), , drop = FALSE])),
      crs = sf::st_crs(points_sf)
    )
  )
}

cog_sf <- cog |>
  XY_to_sf(coords = c("est_x", "est_y"), crs_to = st_crs(pred_cell_sf))

cog_segments_sf <- cog_sf |>
  dplyr::group_split(region, time_period) |>
  purrr::map(make_cog_segments) |>
  purrr::compact() |>
  purrr::reduce(rbind) |>
  filter(region == "all")

prediction_background <- pred_cell_sf |>
  select(-region) |>
  filter(year == last_historical_year | year == 2060)

ggplot() +
  geom_sf(data = pacea::bc_coast, fill = "grey90", linewidth = 0.2) +
  geom_sf(data = prediction_background, aes(fill = density_km2), alpha = 0.4, colour = NA) +
  scale_fill_viridis_c(trans = "log10") +
  geom_sf(data = cog_segments_sf, aes(colour = year), linewidth = 0.8,
        arrow = grid::arrow(length = unit(0.18, "cm"), type = "closed")) +
  # geom_sf(data = cog_sf |> slice_max(year), aes(colour = year), size = 3) +
  scale_colour_viridis_c(option = "magma", end = 0.7) +
  # geom_sf(data = region_borders) +
  labs(colour = "Year", title = "Centre of gravity trajectory, Pacific Spiny Dogfish (WCVI)") +
  gfplot::coord_sf_auto(pred_cell_sf) +
  facet_grid(region ~ time_period)
dir.create("figs")
ggsave("figs/cog-dogfish-eg.png", width = 7, height = 4)

ggplot() +
  geom_sf(data = pacea::bc_coast, fill = "grey90", linewidth = 0.2) +
  geom_sf(data = prediction_background, aes(fill = density_km2), alpha = 0.4, colour = NA) +
  scale_fill_viridis_c(trans = "log10") +
  geom_sf(data = cog_segments_sf, aes(colour = year), linewidth = 0.8,
        arrow = grid::arrow(length = unit(0.18, "cm"), type = "closed")) +
  geom_sf(data = cog_sf |> slice_max(year), aes(colour = year), size = 3) +
  scale_colour_viridis_c(option = "magma", end = 0.7) +
  geom_sf(data = region_borders) +
  labs(colour = "Year", title = "Centre of gravity trajectory, Pacific Spiny Dogfish (WCVI)") +
  gfplot::coord_sf_auto(pred_cell_sf) +
  facet_grid(region ~ time_period)


# These are not plotted right now but we could print them
ew <- ggplot(cog, aes(year, est_x)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_x, ymax = upr_x)) +
  facet_wrap(~ region)

ns <- ggplot(cog, aes(year, est_y)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y)) +
  facet_wrap(~ region)
```

#### Area occupied

Total area containing 90\% of a species' abundance/biomass: describing how spread out a population is and indicating expansion or contraction in regional occupancy through time

Three ways to display it on the shiny app plot:

-[x] Area used (%) (by regions)

-[x] Area used (km2)

-[x] Area Efficiency

```{r}
ranked_accumulation_by_sims <- sims |> # @optimize: may need data.table here
  group_by(time_period, region, year, sim) |>
  # step 1 - rank by density within groups
  arrange(desc(log_density), .by_group = TRUE) |>
  # step 2 - cumulative abundance
  mutate(
    total_abundance = sum(exp(log_abundance)),
    total_area = sum(area),
    cumulative_abundance = cumsum(exp(log_abundance)),
    percent_abundance = cumulative_abundance / total_abundance * 100,
    cumulative_area = cumsum(area),
    id = row_number()
  ) # for plotting accumulation curves

# Accumulation curves -----
# ranked_accumulation_by_sims |>
#   filter(sim %in% 1:10) |>
#   filter(year %in% c(2004, last_historical_year, 2060)) |>
# ggplot(aes(x = id, y = cumulative_abundance)) +
#   geom_point(aes(colour = year)) +
#   facet_wrap(~ region, scales = "free_x")

# ranked_accumulation_by_sims |>
#   filter(sim %in% 1:10) |>
#   filter(year %in% c(2004, last_historical_year, 2060)) |>
# ggplot(aes(x = id, y = cumulative_area)) +
#   geom_point(aes(colour = year)) +
#   facet_wrap(~ region, scales = "free_x")


# Using approx instead for linear interpolation like is done for range edges rather than choosing a value before (as in Halibut paper) or after (as I was doing before) the threshold is crossed
thresholds <- c(50, 75, 90, 95)

quantile_eao_by_sim <- ranked_accumulation_by_sims |>
  reframe(
    threshold = thresholds,
    area_threshold = approx(x = percent_abundance, y = cumulative_area,
                                xout = thresholds, rule = 2, ties = "ordered")$y,
    percent_area_used = area_threshold / dplyr::first(total_area),
    area_efficiency    = area_threshold / dplyr::first(total_abundance)
  ) |>
  dplyr::select(time_period, region, year, sim, threshold, area_threshold, percent_area_used, area_efficiency)


quantile_eao <- quantile_eao_by_sim |>
  summarise_sim_ci(
    cols = c(area_threshold, percent_area_used, area_efficiency),
    by = c(time_period, region, year, threshold)
  )

eao_metric_meta <- tibble::tribble(
  ~metric,             ~name,                  ~units,
  "area_threshold",    "Area occupied",        "km^2^ / year",
  "percent_area_used", "Percent of area used", "logit / year",
  "area_efficiency",    "Area efficiency",        "km^2^ / kg or count"
)

eao_lm <- quantile_eao_by_sim |>
  # transform responses before lm fitting
  mutate(
    centred_year      = year - last_historical_year,
    area_threshold    = area_threshold, # (units = km2)
    percent_area_used = qlogis(percent_area_used), # (units = dimensionless)
    area_efficiency   = area_efficiency # (units = km2 / biomass unit -- kg in dogfish example)
  ) |>
  tidyr::pivot_longer(c(area_threshold, percent_area_used, area_efficiency),
                      names_to = "metric", values_to = "value") |>
  summarise_sim_trend(value = value, by = c(time_period, region, metric, threshold)) |>
  left_join(eao_metric_meta, by = "metric")

eao_summary <- eao_lm |>
  filter(term == "centred_year") |>
transmute(
    indicator = paste0("eao-", metric, "-", threshold),
    name      = paste0(name, " (", threshold, "%)"),
    time_period,
    region,
    estimate,
    lwr,
    upr,
    se,
    prob_direction,
    units
  )

# @Note to self - as a species becomes less patchy (less aggregated), area efficiency should increase because it takes more area to hold X% of the population; dividing the area threshold by abundance standardises this effect from total abundance (i.e., without it, as total abundance increases it obviously takes more area to hold the same amount of individuals)
```

```{r compare-approx-vs-slice1}
#| echo: false
#| eval: false
# step 3 - calculate values at abundance thresholds
# thresholds <- c(50, 75, 90, 95)
# quantile_eao_by_sim <- purrr::map(thresholds, \(t) {
#   #@NOTE - right now this depends on maintaining the grouping of ranked_accumulation_by_sims, maybe we should code this in here/add a check here.
#   ranked_accumulation_by_sims |>
#     filter(percent_abundance >= t) |>
#     slice(1) |> # first cell that crosses, per group; should we be interpolating like is done for range edge? use approx?
#     mutate(
#       threshold         = t,
#       area_threshold    = cumulative_area,
#       percent_area_used = area_threshold / total_area,
#       area_efficiency   = area_threshold / total_abundance
#     )
# }) |>
#   purrr::list_rbind() |>
#   ungroup() |>
#   select(time_period, region, year, sim, threshold,
#          area_threshold, percent_area_used, area_efficiency)
```


```{r plot-area-occupied}
plot_area_metric <- function(df, colour_var, facet_var = NULL,
                             title = NULL,
                             boundary_year = last_historical_year) {
  p <- ggplot(df) +
    aes(x = year, y = est, ymin = lwr, ymax = upr,
        colour = {{ colour_var }}, fill = {{ colour_var }},
        group = {{ colour_var }}) +
    geom_vline(xintercept = boundary_year, linetype = 2, colour = "grey60") +
    geom_line() +
    geom_ribbon(alpha = 0.2, colour = NA) +
    scale_x_continuous(expand = expansion()) +
    labs(y = NULL, title = title)

  if (!rlang::quo_is_null(rlang::enquo(facet_var))) { # enquo captures without evaluating; then can check if quo is null; no need to evaluate/check whether column exists
    p <- p + facet_wrap(vars({{ facet_var }}), ncol = 1, scales = "free_y")
  }
  p
}

quantile_eao_long <- quantile_eao |>
  tidyr::pivot_longer(
    cols = -c(time_period, region, year, threshold),
    names_to = c(".value", "metric"),
    names_pattern = "(est|lwr|upr|se)_(.*)"
  )

# One plot per metric (different y-units). Filter to the metric HERE,
# not inside the function. pmap over the meta table so title = the
# human-readable name and units are available if you want them.
area_plots <- purrr::pmap(eao_metric_meta, \(metric, name, units) {
  quantile_eao_long |>
    filter(region != "all") |> # only look at subregions
    filter(.data$metric == .env$metric, threshold == 90) |>
    plot_area_metric(
      colour_var = region,
      facet_var = NULL,
      title = name
    ) +
    scale_colour_viridis_d(end = 0.7) +      # region is discrete
    scale_fill_viridis_d(end = 0.7)
})
```

```{r plot-area-occupied-composite}
#| fig-width: 7
#| fig-height: 10
patchwork::wrap_plots(area_plots, ncol = 1) +
  patchwork::plot_layout(guides = "collect")
```

```{r plot-area-occupied-individual}
#| eval: false
#| echo: false
area_plots[[1]]
area_plots[[2]]
area_plots[[3]]
```


#### Range edge

* Change in the position of the southernmost (trailing, 5% quantile) range edge and northernmost (leading, 95% quantile) range edge of modeled species' distribution

* Trailing edge (5% quantile): the latitude beyond which (southward) no more than

  5% of the population's (density-weighted) abundance extends.

* Leading edge (95% quantile): the latitude beyond which (northward) no more than
  5% of the population's (density-weighted) abundance extends.

* What is this really meaning when we have only a small part of the distribution??

* Not calculated by region


$$\frac{\sum_{i \le k} d_i A}{\sum_i d_i A} = \frac{A\sum_{i \le k} d_i}{A\sum_i d_i} = \frac{\sum_{i \le k} d_i}{\sum_i d_i}$$
But we may have unequal area grid cells - e.g., the HBLL? Though realistically this might be the only case where this actually happens.

So we need:

$$\frac{\sum_{i \le k} d_i A}{\sum_i d_i A}$$

```{r range-edges}
# step 1 - order spatial locations along specified axis (Y = N/S (N > S); X = E/W (W > E))
# step 2 - calculate cumulative proportion of total density (for unequal area - abundance) along that axis
# step 3 - Find positions where cumulative proportion equals target quantities (5%, 95%)
# step 4 - Summarise from simulations for uncertainty

# @Question - do we want/need the EW range?
# @Question - does the distance along coast make more sense? e.g., see Fredston et al. 2021

range_edge_quantiles <- c(trailing = 0.05, leading = 0.95)

range_edge_NS_by_sim <- sims |>
  filter(region == "all") |>
  get_range_edge_sim(
    rank_var = Y,
    quantiles = range_edge_quantiles,
    by = c(time_period, region, year, sim) # including region keeps the column even though we only use "all"
  )
range_edge_EW_by_sim <- sims |>
  filter(region == "all") |>
  get_range_edge_sim(
    rank_var = X,
    quantiles = range_edge_quantiles,
    by = c(time_period, region, year, sim)
  )

range_edge_NS <- range_edge_NS_by_sim |>
  summarise_sim_ci(Y_edge, c(time_period, region, year, edge))
range_edge_EW <- range_edge_EW_by_sim |>
  summarise_sim_ci(X_edge, c(time_period, region, year, edge))

range_NS_lm <- range_edge_NS_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  summarise_sim_trend(value = Y_edge, by = c(time_period, region, edge))
range_EW_lm <- range_edge_EW_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  summarise_sim_trend(value = X_edge, by = c(time_period, region, edge))

edge_meta <- tibble::tribble(
  ~axis, ~edge,      ~indicator,   ~name,
  "NS",  "leading",  "edge-north", "Edge Shift (N)",
  "NS",  "trailing", "edge-south", "Edge Shift (S)",
  "EW",  "leading",  "edge-west",  "Edge Shift (W)", # do these names change depending on coast?
  "EW",  "trailing", "edge-east",  "Edge Shift (E)"
)

edge_summary <- bind_rows(
  range_NS_lm |> mutate(axis = "NS"),
  range_EW_lm |> mutate(axis = "EW")
) |>
  filter(term == "centred_year") |>
  left_join(edge_meta, by = c("axis", "edge")) |>
  transmute(
    indicator,
    name,
    time_period,
    region,
    estimate,
    lwr,
    upr,
    se,
    prob_direction,
    units = "km / year"
  )

```

```{r approx-scratch}
#| eval: false
#| echo: false
# @Note: no zero-abundance failsafe here, unlike sdmTMB's `if (total_dens > 0)` guard: https://github.com/pbs-assess/sdmTMB/blob/891d8e7f914759338ef91d078032ee85d2ed778b/R/get-range-edge.R#L201-L202
# For `approx()`: `rule` controls out-of-range xout (a quantile outside [min(cumul_prop), max(cumul_prop)]):
#   rule = 1 -> NA           (refuse to extrapolate; edge is off-grid)
#   rule = 2 -> nearest end  (clamp to the boundary cell's Y) --> what sdmTMB does

# # @Notes - me understanding how approx maps to the get-range-edge.R code L. 211 - 231
# # approx example:
# cumul_prop     <- c(0.05, 0.20, 0.45, 0.70, 0.90, 1.00)   # x: the CDF
# axis_ordered <- c(5300, 5302, 5304, 5306, 5308, 5310)   # y: northing
# q <- 0.99                                                # target proportion
# cross_idx <- which(cumul_prop >= q)[1]
# idx1 <- cross_idx - 1
# idx2 <- cross_idx

# prop1 <- cumul_prop[idx1]
# prop2 <- cumul_prop[idx2]
# axis1 <- axis_ordered[idx1]
# axis2 <- axis_ordered[idx2]

# # Interpolate
# weight <- (q - prop1) / (prop2 - prop1) # oh, this says q is a third of the way between prop1 and prop2
# edge_pos <- axis1 + weight * (axis2 - axis1) # add 1/3 of the difference between the cross id axis values
# edge_pos
# approx(cumul_prop, axis_ordered, xout = q, rule = 2)$y

# Understanding the approx rules
# q = 0.02  | rule=1:     NA | rule=2:   5300 | sdmTMB:   5300   <- BELOW range: southernmost cell alone already holds >2% -> trailing edge is off-grid
# q = 0.05  | rule=1:   5300 | rule=2:   5300 | sdmTMB:   5300   <- exactly on the first point
# q = 0.50  | rule=1: 5304.4 | rule=2: 5304.4 | sdmTMB: 5304.4   <- interior (interpolation); rule irrelevant
# q = 0.975 | rule=1: 5309.5 | rule=2: 5309.5 | sdmTMB: 5309.5   <- interior; upper quantiles are ALWAYS interior since cum_prop reaches 1.0
# q = 1.00  | rule=1:   5310 | rule=2:   5310 | sdmTMB:   5310   <- exactly on the last point (== max), so NOT out of range -> not NA
# q > 1.00  | rule=1:     NA | rule=2:   5310 | sdmTMB:     NA   <- only here does rule=1 give NA; but this doens't apply because we are looking at quantiles (so q <=1 )

```


#### Average depth

* Density-weighted average depth (50th percentile). Indicates changes in depth niche over time

```{r mean-depth}
# @Question - mean depth, not mean log_depth correct?
d_by_sim <- sims |>
  group_by(time_period, region, year, sim) |>
  summarise(
    mean_depth = weighted.mean(depth_m, w = exp(log_abundance)),
    .groups = "drop"
  )

mean_depth_sims <- d_by_sim |>
  summarise_sim_ci(mean_depth, c(time_period, region, year))

ggplot(data = mean_depth_sims) +
  aes(x = year, y = est, ymin = lwr, ymax = upr, groups = region) +
  geom_line() +
  geom_ribbon(alpha = 0.3) +
  scale_y_reverse() +
  labs(x = "Year", y = "Mean depth") +
  facet_wrap(~ region)

depth_lm <- d_by_sim |>
  mutate(centred_year = year - last_historical_year) |>
  summarise_sim_trend(value = mean_depth, by = c(time_period, region))

depth_summary <- depth_lm |>
  filter(term == "centred_year") |>
  transmute(
    indicator = "depth",
    name = "Depth Trend",
    time_period,
    region,
    estimate,
    lwr,
    upr,
    se,
    prob_direction,
    units = "m / year"
  )
```

#### Outputs for shiny app indicator wheel

Note - these are results from simple linear models - the response has not been scaled with `scale()`.

```{r, results=TRUE}

# @Question: should abundance/biomass trend be reported as % change / year instead of the raw log-scale slope?
bind_rows(
  abundance_summary,
  prop_abundance_summary,
  eao_summary |> filter(grepl("90", indicator)), # use 90% threshold for now
  cog_summary,
  edge_summary,
  depth_summary) |>
  select(name, time_period, region, estimate, se, units) |>
  mutate(across(c(estimate, se), ~ signif(.x, 3))) |>
  knitr::kable(
    col.names = c("Indicator", "Time period", "Region", "Estimate", "SE", "Units"), # option to add prob_value
    align = "llllrrl",
    booktabs = TRUE
  ) |>
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8) |>
  kableExtra::column_spec(2, width = "4cm") |>
  kableExtra::column_spec(3, width = "1.5cm") |>
  kableExtra::column_spec(4, width = "1.3cm") |>
  kableExtra::column_spec(5, width = "1.7cm") |>
  kableExtra::column_spec(6, width = "1.7cm") |>
  kableExtra::column_spec(7, width = "2.2cm")

```
